Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2013-01-08 paper.tex
more of Vaughan's edits

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
This problem does not occur with the least square estimator.

@@

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\hat{\theta} - \theta_0\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\hat{\theta} - \theta_0\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Let the noise sequence $\{w_i\}$ be complex Gaussian with independent real and imaginary parts having zero mean and variance $\sigma^2$.  The joint density function of the real and imaginary parts is
\[
\frac{1}{2\pi\sigma^2}e^{-\frac{1}{2\sigma^2}(x^2 + y^2)}.
\]
Theorems~\ref{thm:consistency}~and~\ref{thm:normality} hold, and since the distribution of $w_1$ is circularly symmetric, the distribution of $R_1e^{j\Phi_1}$ is identical to the distribution of $1 + \frac{1}{\rho_0} w_1$.
%and the joint density function of the real and imaginary parts of $Z$ is
%\[
%p_Z(x,y) = \rho_0^2 p( \rho_0 (x - 1), \rho_0 y ) = \frac{\kappa^2}{2\pi}e^{-\kappa^2(x ^2 - 2x + 1 + y^2) }
%\] 
%where $\kappa = \tfrac{\rho_0}{\sigma}$.  
It can be shown that
\[
%g(\phi) = \frac{\cos^2(\phi)}{4\sqrt{\pi}}\left( \frac{e^{-\kappa^2} }{\sqrt{\pi}} + \kappa \Psi(\phi)  e^{-\kappa^2\sin^2(\phi)}\cos(\phi) \right)
g(\phi) = \frac{\cos\phi}{2\pi}e^{-\frac{1}{2}\kappa^2} + \frac{\Psi(\kappa \cos\phi)}{\kappa\sqrt{2\pi}}  e^{-\frac{1}{2}\kappa^2\sin^2\phi} \big(1 + \kappa^2\cos^2\phi \big)
\]
where $\kappa = \rho_0/\sigma$ and $\Psi(x) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^{x} e^{-x^2/2} dx$
is the cumulative density function of the standard normal.
%\[
%\Psi(\phi) = \frac{1}{2} + \frac{1}{2} \erf\left(\frac{\kappa \cos\phi}{\sqrt{2}}\right)
%\]
%and $\erf(x) = \frac{2}{\sqrt{\pi}}\int_0^{x} e^{-t^2} dt$ is the error function.  
The value of $A_1, A_2, B_1$ and $B_2$ can be efficiently computed by numerical integration using these formula.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilots symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ (i.e. coherent detection) the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality} (and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  The prediction is made by dividing the asymptotic covariance matrix in Theorem~\ref{thm:normality} by $L$.  The theorem accurately predicts the behaviour of the phase estimator when $L$ is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$ and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Theorem~\ref{thm:normality} does not model this behaviour in the sense that for any fixed $L$ there exists a sufficiently small value of $SNR$ such that Theorem~\ref{thm:normality} does not produce accurate predictions.

@@

@@

We showed, under some conditions on the distribution of the noise, that the phase estimator $\hat{\theta}$ is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho}_0$ is biased, and converges to $G(0)\rho_0$.  This bias is large when the signal to noise ratio is small.  It would be interesting to investigate methods for correcting this bias.  A method for estimating $G(0)$ at the receiver appears to be required.

Monte Carlo simulations were used to assess the performance of the least squares estimator and also to validate our asymptotic theory.  Interestingly, when the SNR is small, it is counterproductive to use the data symbols to estimate the phase (Figure~\ref{fig:plotphaseQPSKmultL}).  This suggests the use of a weighed objective function, which would be an interesting topic for future research.

>>>>

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\begin{equation}\label{eq:viterbiviterbi}
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\end{equation}
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
This problem does not occur with the least square estimator.

@@

The paper is organised in the following way.  Section~\ref{sec:least-squar-estim} extends Mackenthun's algorithm for the coherent case, when both pilot symbols and data symbols exist.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} states two theorems that describe the statistical properties of the least squares estimator of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some reasonably general assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\hat{\theta} - \theta_0\rangle_\pi$ converges almost surely to zero and that $\sqrt{L}\langle\hat{\theta} - \theta_0\rangle_\pi$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  Sections~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm} provide proofs of the theorems stated in Section~\ref{sec:stat-prop-least}.  In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our expressions for the asymptotic distribution can be simplified.  Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Let the noise sequence $\{w_i\}$ be complex Gaussian with independent real and imaginary parts having zero mean and variance $\sigma^2$.  The joint density function of the real and imaginary parts is
\[
\frac{1}{2\pi\sigma^2}e^{-\frac{1}{2\sigma^2}(x^2 + y^2)}.
\]
Theorems~\ref{thm:consistency}~and~\ref{thm:normality} hold, and since the distribution of $w_1$ is circularly symmetric, the distribution of $R_1e^{j\Phi_1}$ is identical to the distribution of $1 + \frac{1}{\rho_0} w_1$.
%and the joint density function of the real and imaginary parts of $Z$ is
%\[
%p_Z(x,y) = \rho_0^2 p( \rho_0 (x - 1), \rho_0 y ) = \frac{\kappa^2}{2\pi}e^{-\kappa^2(x ^2 - 2x + 1 + y^2) }
%\] 
%where $\kappa = \tfrac{\rho_0}{\sigma}$.  
It can be shown that
\[
%g(\phi) = \frac{\cos^2(\phi)}{4\sqrt{\pi}}\left( \frac{e^{-\kappa^2} }{\sqrt{\pi}} + \kappa \Psi(\phi)  e^{-\kappa^2\sin^2(\phi)}\cos(\phi) \right)
g(\phi) = \frac{\cos\phi}{2\pi}e^{-\frac{1}{2}\kappa^2} + \frac{\Psi(\kappa \cos\phi)}{\kappa\sqrt{2\pi}}  e^{-\frac{1}{2}\kappa^2\sin^2\phi} \big(1 + \kappa^2\cos^2\phi \big)
%g(\phi) = \frac{\cos\phi}{2\pi}e^{-\kappa} + \frac{\Psi(\kappa \cos\phi)}{\kappa\sqrt{2\pi}}  e^{-\kappa\sin^2\phi} \big(1 + 2\kappa\cos^2\phi \big)
\]
where $\kappa = \rho_0/\sigma$ and $\Psi(x) = \frac{1}{\sqrt{2\pi}} \int_{-\infty}^{x} e^{-x^2/2} dx$
is the cumulative density function of the standard normal.
%\[
%\Psi(\phi) = \frac{1}{2} + \frac{1}{2} \erf\left(\frac{\kappa \cos\phi}{\sqrt{2}}\right)
%\]
%and $\erf(x) = \frac{2}{\sqrt{\pi}}\int_0^{x} e^{-t^2} dt$ is the error function.  
The value of $A_1, A_2, B_1$ and $B_2$ can be efficiently computed by numerical integration using this formula.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} show the sample mean square error (MSE) of the phase estimator when $M=2,4,8$ with $L=4096$ and for varying proportions of pilots symbols $\abs{P} = 0, \tfrac{L}{32}, \tfrac{L}{8}, \tfrac{L}{2}, L$.  When $\abs{P} \neq 0$ (i.e. coherent detection) the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}_\pi^2$.  Otherwise, when $\abs{P}=0$ the mean square error is computed as $\tfrac{1}{T}\sum_{i=1}^T\sfracpart{\hat{\theta}_i - \theta_0}^2$ as in~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  The dots, squares, circles and crosses are the results of Monte-Carlo simulations with the least square estimator.  The solid lines are the estimator MSEs predicted by Theorem~\ref{thm:normality} (and by Theorem~2 from~\cite{McKilliam_leastsqPSKnoncoICASSP_2012} when $\abs{P} = 0$).  The prediction is made by dividing the asymptotic covariance matrix by $L$.  The theorem accurately predicts the behaviour of the phase estimator when $L$ is sufficiently large.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$ when $\abs{P} \neq 0$ and the uniform distribution on $[-\tfrac{\pi}{M}, \tfrac{\pi}{M})$ when $\abs{P}=0$~\cite{McKilliam_leastsqPSKnoncoICASSP_2012}.  Theorem~\ref{thm:normality} does not model this behaviour in the sense that for any fixed $L$ there exist sufficiently small values of SNR for which Theorem~\ref{thm:normality} does not produce accurate predictions of the MSE.

The figures display some interesing behaviour.  When the SNR is small \dots.  When the SNR is large \dots.

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} also display the sample MSE of the noncoherent phase estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983} described by~\eqref{eq:viterbiviterbi}.  This estimator requires selection of a function $F$ that transforms the amplitude of each sample prior to the final estimation step.  % (see equation (2) in~\cite{ViterbiViterbi_phase_est_1983}).  
Viterbi and Viterbi propose several viable alternatives, from which we have chosen $F(x) = 1$.  The Viterbi and Viterbi estimator is only applicable in the noncoherent setting, i.e. when $\abs{P} = 0$.  The sample MSE of the least squares estimator (when $\abs{P} = 0$) and the Viterbi and Viterbi estimator is similar.  The least squares estimator appears slightly more accurate for some values of SNR.

@@

\section{When the SNR is small}

In this section we study the behaviour of the least squares estimator when SNR goes to zero.  We consider only the case when the noise samples $w_1,\dots,w_L$ are independent and identically distributed circularly symmetric and Gaussian with real and imaginary parts having variance $\sigma^2$.  To simplify our notation denote the signal to noise ratio by $\nu = \frac{\rho_0^2}{2\sigma^2}$.  In this case the function $g(\phi)$ takes to form
\[
g(\phi) = \frac{1}{2\sqrt{\pi\nu}} + \frac{\cos(\phi)}{2\pi} + o(1)
\]
as $\nu \rightarrow 0$.  Now
\[
A_1 = \int_{-\pi}^{\pi} g(\phi) \sin^2(\phi) d\phi = \frac{\pi}{2\sqrt{\pi\nu}} + O(1),
\]
\begin{align*}
A_2 &= \int_{-\pi}^{\pi} g(\phi) \sin^2\fracpart{\phi} d\phi \\
&= \frac{M}{2\sqrt{\pi\nu}} \int_{-\pi/M}^{\pi/M}\sin^2(\phi) d\phi + O(1) \\
&= \frac{\pi - M\sin(\tfrac{\pi}{M})}{2\sqrt{\pi\nu}},
\end{align*}
\begin{align*}
h_2(0) &= \int_{-\pi}^{\pi} g(\phi) \cos\fracpart{\phi} d\phi \\
&= \frac{M}{2\sqrt{\pi\nu}} \int_{-\pi/M}^{\pi/M}\cos(\phi) d\phi + O(1) \\
&= \frac{M\sin(\tfrac{\pi}{M})}{\sqrt{\pi\nu}}
\end{align*}
and so
\begin{align*}
H &= h_2(0) - 2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}) \\
&= h_2(0) -  \frac{M\sin(\pi/M)}{\sqrt{\pi\nu}}\\
&= O(1)
\end{align*}
Ok, you need to expand to one more term.
\section{When the SNR is large}

@@

We showed, under some reasonably general conditions on the distribution of the noise, that the phase estimator $\hat{\theta}$ is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho}_0$ is biased, and converges to $G(0)\rho_0$.  This bias is large when the signal to noise ratio is small.  It would be interesting to investigate methods for correcting this bias.  A method for estimating $G(0)$ at the receiver appears to be required.

Monte Carlo simulations were used to assess the performance of the least squares estimator and also to validate our asymptotic theory.  Interestingly, when the SNR is small, it is counterproductive to use the data symbols to estimate the phase (Figure~\ref{fig:plotphaseQPSKmultL}).  This suggests the use of a weighted objective function, which would be an interesting topic for future research.
